Mechanical loading prediction though accelerometry data during walking and running

### 2021-05-12 manuscript/manuscript.Rmd
Adapt to the new results

The ground reaction force (GRF) is arguably the most studied mechanical loading variable, and its relationship with the bone adaptations to exercise is well established [@Turner2003; @Rantalainen2008; @Rantalainen2009]. Due to its importance, some studies have already attempted to develop peak GRF (pGRF) prediction equations based on accelerometry data in several populations, such as children [@Neugebauer2012], adults [@Fortune2014; @Neugebauer2014], patients with obesity [@Veras2020] and military [@Neugebauer2018], but with the majority of these studies including only activities with eliciting a low loading magnitude. Besides loading magnitude, loading rate (LR) is another variable related with bone health [@Rowlands2014] that could potentially be predicted by accelerometer data, but with apparently lower prediction accuracy compared with pGRF [@Fortune2014; @Veras2020]. Also, to the best of our knowledge, peak LR (pLR) prediction through accelerometer data has only been tested during walking.

@@

Data collection took place at Porto Biomechanics Laboratory (LABIOMEP-UP) on an instrumented treadmill with built-in FP (AMTI Corporation, Watertown, USA). Before starting the data collection, participants were given time to warm up on the treadmill. The protocol consisted of walking and running in 13 incremental speeds, with 0º inclination, from 2 to 14 km·h^-1^ with 1 km·h^-1^ increment. Walking and running speeds ranged from 2 to 6 km·h^-1^ and from 7 to 14 km·h^-1^, respectively. Each walking speed lasted 1 minute and each running speed 30 seconds. To ensure single foot contact on each FP during the trial, feedback regarding location on the treadmill was provided by a researcher. Participants wore their own sports shoes throughout the trials.

@@

To assess the models prediction accuracy, mean absolute error (MAE), mean absolute percent error (MAPE) and root mean square error (RMSE) were calculated. Although there is no standard threshold defining an acceptable error for mechanical loading prediction, based on previous studies we considered a MAPE <8\% to be an accurate prediction [@Fortune2014; @Neugebauer2014; @Ancillao2018]. Also, methods agreement was evaluated with Bland-Altman plots by plotting the difference between the actual (FP measured) and predicted values against their mean. Bias was computed as the mean of these differences and the limits of agreement were obtained using ±1.96 standard deviation of the mean between actual and predicted values [@Bland1986]. In order to test whether there was a consistent over or underestimation of the actual values, one-sample *t* tests were performed to assess if the difference between actual and predicted values were different than zero. Linear regressions were applied to identify if the magnitude of the mean between actual and predicted values influenced the magnitude of their difference. This was done to ascertain if the prediction accuracy was constant throughout the assayed magnitude range, which is known as proportional bias [@Giavarina2015]. These analyses were conducted separately for data from each accelerometer placement and outcome.

@@

Our recruited sample presented a broad range of body masses, from normal weight to severe obesity. It would be expected that participants with higher body mass index (BMI) values, due to low motor ability and physical capacity, are able to perform the trials only at the lowest speeds. This was precisely what happened and, as can be observed in Figure 1, from 2 to 5 km·h^-1^ the BMI median was around 30 kg·m^-2^, decreasing to around 27 kg·m^-2^ at 6 km·h^-1^, which was the fastest walking speed, with a further reduction to less than 25 kg·m^-2^ at the running speeds. Furthermore, only one participant with a BMI higher than 30 kg·m^-2^ was able to run. This implies in a sample size reduction from `r filter(sample_size, acc_placement == "hip" & vector == "resultant" & speed == "2")$pGRF_N` at the slowest speed to `r filter(sample_size, acc_placement == "hip" & vector == "resultant" & speed == "14")$pGRF_N` at the fastest.

>>>>

The ground reaction force (GRF) is arguably the most studied mechanical loading variable, and its relationship with the bone adaptations to exercise is well established [@Turner2003; @Rantalainen2008; @Rantalainen2009]. Due to its importance, some studies have already attempted to develop peak GRF (pGRF) prediction equations based on accelerometry data in several populations, such as children [@Neugebauer2012], adults [@Fortune2014; @Neugebauer2014], patients with obesity [@Veras2020] and military [@Neugebauer2018], but with the majority of these studies including only activities eliciting a low loading magnitude. Besides loading magnitude, loading rate (LR) is another variable related with bone health [@Rowlands2014] that could potentially be predicted by accelerometer data, but with apparently lower prediction accuracy compared with pGRF [@Fortune2014; @Veras2020]. Also, to the best of our knowledge, peak LR (pLR) prediction through accelerometer data has only been tested during walking.

@@

Data collection took place at Porto Biomechanics Laboratory (LABIOMEP-UP) on an instrumented front-back split-belt treadmill with built-in FP (AMTI Corporation, Watertown, USA). Before starting the data collection, participants were given time to familiarize with the treadmill. The protocol consisted of walking and running in 13 incremental speeds, with 0º inclination, from 2 to 14 km·h^-1^ with 1 km·h^-1^ increment. Walking and running speeds ranged from 2 to 6 km·h^-1^ and from 7 to 14 km·h^-1^, respectively. Each walking speed lasted 1 minute and each running speed 30 seconds. To ensure single foot contact on each FP during the trial, feedback regarding location on the treadmill was provided by a researcher. Participants wore their own sports shoes throughout the trials.

@@

To assess the models prediction accuracy, mean absolute error (MAE), mean absolute percent error (MAPE) and root mean square error (RMSE) were calculated. Although there is no standard threshold defining an acceptable error for mechanical loading prediction, based on previous studies we considered a MAPE <8\% to be an accurate prediction [@Fortune2014; @Neugebauer2014; @Ancillao2018]. Also, methods agreement was evaluated with Bland-Altman plots by plotting the difference between the actual (FP measured) and predicted values against their mean. Bias was computed as the mean of these differences and the limits of agreement were obtained using ±1.96 standard deviation of the bias [@Bland1986]. In order to test whether there was a consistent over or underestimation of the actual values, one-sample *t* tests were performed to assess if the difference between actual and predicted values were different than zero. Linear regressions were applied to identify if the magnitude of the mean between actual and predicted values influenced the magnitude of their difference. This was done to ascertain if the prediction accuracy was constant throughout the assayed magnitude range, which is known as proportional bias [@Giavarina2015]. These analyses were conducted separately for data from each accelerometer placement and outcome.

@@

Our recruited sample presented a broad range of body masses, from normal weight to severe obesity. It would be expected that participants with higher body mass index (BMI) values, due to low motor ability and physical capacity, are able to perform the trials only at the lowest speeds. This was precisely what happened and, as can be observed in Figure 1, the number of data points corresponding to subjects with BMI > 30 kg·m^-2^ greatly decreases in the speeds above 6 km·h^-1^.
